Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-08-07 analysis/paper/paper.qmd
updates

df <- tibble(
  material = c("Lava", "Dolomite", "Chert", "Other"),
  spheroids = c(9, 2, 0, 1),
  cores = c(62, 21, 4, 7),
  total = c(637, 209, 53, 56),
)

df_long <- df %>%
  pivot_longer(cols = -material, names_to = "category", values_to = "count") %>%
  group_by(category) %>%
  mutate(prop = count / sum(count) * 100) %>%
  ungroup() %>%
  mutate(
    category = factor(category, levels = c("spheroids", "cores", "total")),
    material = factor(material, levels = c("Lava", "Dolomite", "Chert", "Other"))
  )

p_raw_material <- 
ggplot(df_long %>% 
         mutate(category = case_when(
           category == "cores" ~ "Core",
           category == "spheroids" ~ "Spheroid",
           category == "total" ~ "Total"
           )), 
       aes(x = material, 
           y = prop, 
           fill = category)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    x = "",
    y = "Percentage (%)",
    fill = "Technological Type"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    scale_fill_manual(values = c(
    "Core" = "#66c2a5", 
    "Spheroid" = "orange", 
    "Total" = "lightgray"
  )) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = c(0.95, 0.95),              
    legend.justification = c("right", "top"),      
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
    legend.box.background = element_rect(color = "black")  
  )

@@

lava_spheroids_text <- df_long %>% filter(material == "Lava", category == "spheroids") %>% pull(text)
dolomite_spheroids_text <- df_long %>% filter(material == "Dolomite", category == "spheroids") %>% pull(text)
chert_spheroids_text <- df_long %>% filter(material == "Chert", category == "spheroids") %>% pull(text)
other_spheroids_text <- df_long %>% filter(material == "Other", category == "spheroids") %>% pull(text)
```

@@

Raw materials at Qianshangying were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others. The overall assemblage is dominated by lava ( `r lava_total_text` ), followed by siliceous dolomite ( `r dolomite_total_text` ), with smaller proportions of chert ( `r chert_total_text` ) and other materials ( `r other_total_text` ). The composition of cores closely mirrors this distribution, comprising `r round(df_long$prop[df_long$material == "Lava" & df_long$category == "cores"],1)`% lava (n = `r df_long$count[df_long$material == "Lava" & df_long$category == "cores"]`), `r round(df_long$prop[df_long$material == "Dolomite" & df_long$category == "cores"],1)`% siliceous dolomite (n = `r df_long$count[df_long$material == "Dolomite" & df_long$category == "cores"]`), `r round(df_long$prop[df_long$material == "Chert" & df_long$category == "cores"],1)`% chert (n = `r df_long$count[df_long$material == "Chert" & df_long$category == "cores"]`), and `r round(df_long$prop[df_long$material == "Other" & df_long$category == "cores"],1)`% other (n = `r df_long$count[df_long$material == "Other" & df_long$category == "cores"]`). In contrast, spheroids exhibit a marked preference for lava (`r lava_spheroids_text`), with reduced use of siliceous dolomite ( `r dolomite_spheroids_text` ) and no use of chert. Compared to both the total assemblage and the cores, spheroids show a stronger reliance on lava and a diminished use of more fragile materials, suggesting a selective strategy favoring raw materials with more stable and homogeneous physical properties.

@@

>>>>

df <- tibble(
  material = c("Lava", "Dolomite", "Chert", "Other"),
  spheroids = c(9, 2, 0, 1),
  cores = c(62, 21, 4, 7),
  Polys_and_Multi_cores = c(26, 6, 3, 2),
  total = c(637, 209, 53, 56),
)

df_long <- df %>%
  pivot_longer(cols = -material, names_to = "category", values_to = "count") %>%
  group_by(category) %>%
  mutate(prop = count / sum(count) * 100) %>%
  ungroup() %>%
  mutate(
    category = factor(category, levels = c("spheroids","Polys_and_Multi_cores", "cores", "total")),
    material = factor(material, levels = c("Lava", "Dolomite", "Chert", "Other"))
  )

p_raw_material <- 
ggplot(df_long %>% 
         mutate(category = case_when(
           category == "cores" ~ "Core",
           category == "spheroids" ~ "Spheroid",
           category == "total" ~ "Total",
           category == "Polys_and_Multi_cores" ~ "Polys and Multi cores"
           )), 
       aes(x = material, 
           y = prop, 
           fill = category)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    x = "",
    y = "Percentage (%)",
    fill = "Technological Type"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    scale_fill_manual(values = c(
    "Core" = "#66c2a5", 
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "lightgray",
    "Total" = "gray"
  )) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = c(0.95, 0.95),              
    legend.justification = c("right", "top"),      
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
    legend.box.background = element_rect(color = "black")  
  )
p_raw_material

@@

lava_spheroids_text <- df_long %>% filter(material == "Lava", category == "spheroids") %>% pull(text)
dolomite_spheroids_text <- df_long %>% filter(material == "Dolomite", category == "spheroids") %>% pull(text)
chert_spheroids_text <- df_long %>% filter(material == "Chert", category == "spheroids") %>% pull(text)
other_spheroids_text <- df_long %>% filter(material == "Other", category == "spheroids") %>% pull(text)

lava_mluti_text <- df_long %>% filter(material == "Lava", category == "Polys_and_Multi_cores") %>% pull(text)
dolomite_mluti_text <- df_long %>% filter(material == "Dolomite", category == "Polys_and_Multi_cores") %>% pull(text)
chert_mluti_text <- df_long %>% filter(material == "Chert", category == "Polys_and_Multi_cores") %>% pull(text)
other_mluti_text <- df_long %>% filter(material == "Other", category == "Polys_and_Multi_cores") %>% pull(text)
```

@@

Raw materials at Qianshangying were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others. The overall assemblage is dominated by lava ( `r lava_total_text` ), followed by siliceous dolomite ( `r dolomite_total_text` ), with smaller proportions of chert ( `r chert_total_text` ) and other materials ( `r other_total_text` ). The composition of cores closely mirrors this distribution, comprising `r round(df_long$prop[df_long$material == "Lava" & df_long$category == "cores"],1)`% lava (n = `r df_long$count[df_long$material == "Lava" & df_long$category == "cores"]`), `r round(df_long$prop[df_long$material == "Dolomite" & df_long$category == "cores"],1)`% siliceous dolomite (n = `r df_long$count[df_long$material == "Dolomite" & df_long$category == "cores"]`), `r round(df_long$prop[df_long$material == "Chert" & df_long$category == "cores"],1)`% chert (n = `r df_long$count[df_long$material == "Chert" & df_long$category == "cores"]`), and `r round(df_long$prop[df_long$material == "Other" & df_long$category == "cores"],1)`% other (n = `r df_long$count[df_long$material == "Other" & df_long$category == "cores"]`). For polyhedrons and multifacial cores, a preference of chert ( `r chert_multi_text` ) and lava ( `r lava_multi_text` ) was shown. In contrast, spheroids exhibit a marked preference for lava (`r lava_spheroids_text`), with reduced use of siliceous dolomite ( `r dolomite_spheroids_text` ) and no use of chert. Compared to both the total assemblage and the cores, spheroids show a stronger reliance on lava and a diminished use of more fragile materials, suggesting a selective strategy favoring raw materials with more stable and homogeneous physical properties.

@@

```{r, echo=FALSE}
library(here)
# 读取无列名的csv
library(here)
data_dir <- here("analysis", "data", "derived_data", "power_spectrum")
csv_files <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE)

# 读取所有 CSV 文件，并加上文件名
all_data <- map_dfr(csv_files, function(file) {
  df <- read.csv(file, header = FALSE)
  colnames(df) <- c("degree", "total_power", "max_amplitude", "total_amplitude")
  df$file <- basename(file)
  return(df)
})

# 添加 group 列（可以多标签：Spheroid, Polyhedron, All）
all_data <- all_data %>%
  mutate(
    group = case_when(
      str_detect(file, regex("Polyhedron|Multifacial", ignore_case = TRUE)) ~ "Polyhedron/Multifacial",
      str_detect(file, regex("Spheroid|Subspheroid", ignore_case = TRUE)) ~ "Spheroid/Subspheroid",
      TRUE ~ NA_character_
    )
  )

# 所有样本一组（不管名字）
all_data_all <- all_data %>%
  mutate(group = "All specimens")

# 合并所有分组
combined_data <- bind_rows(all_data, all_data_all) %>%
  filter(!is.na(group))

# 计算每组的 degree 平均值和标准差
summary_df <- combined_data %>%
  group_by(group, degree) %>%
  summarise(
    mean_power = mean(total_power, na.rm = TRUE),
    sd_power = sd(total_power, na.rm = TRUE),
    .groups = "drop"
  )

# 绘图，带误差带
ggplot(summary_df_filtered, aes(x = degree, y = mean_power, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(
    ymin = pmax(mean_power - sd_power, 1e-10),
    ymax = mean_power + sd_power
  ), alpha = 0.2, linetype = 0) +
  scale_y_log10(
  breaks = c(1e-8, 1e-6, 1e-4, 1e-2, 1),
  labels = parse(text = c("10^-8", "10^-6", "10^-4", "10^-2", "10^0"))
  )+
  scale_x_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 4),
    expand = c(0, 0)
  ) +
  coord_cartesian(clip = "on") +
  scale_color_manual(values = c(
    "All specimens" = "#66c2a5",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "lightgray"
  )) +
  scale_fill_manual(values = c(
    "All specimens" = "#66c2a5",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "lightgray"
  )) +
  labs(
    title = "Mean Power Spectrum with Standard Deviation",
    x = "Degree (l)",
    y = "Mean Total Power (log scale)",
    color = "Group",
    fill = "Group"
  ) +
  theme_minimal() +
  theme(plot.margin = margin(0, 0, 0, 0))
```
